---
title: "Tidy Models"
author: "Anwar Dahir"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 


```{r, warning=FALSE, message=FALSE, include=FALSE}
library('tidyverse') ; library('modeldata')
```


## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.

```{r load data, include = FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?

In this report, we examine what factors predict a cricket's chirp rate.


```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```


## Temperature affects chirp rate

Based on a scatter plot of temperature and chirping, it seems that as temperature increases, the rate of chirping also increases.

```{r temp, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping


```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
  
species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```


The graph shows that for both type of species there is a positive relationship between temperature and chirp rate. The model summary shows that both the predictors are statistically significant and the R square for the model is 0.9896.

## Interactions

```{r species historgram, echo = FALSE}
ggplot(crickets, aes(x = rate, fill = species)) +
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```
```{r interactions, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets) 
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
```
The data is checked for ineractions but finally, it is decided to stay wit the original model.


# Diamonds Data Analysis

The scatter plot shows that the carat and price has a positive relationship while the boxplot shows that the average price is highest for premium cut.

```{r, message=FALSE}
diamonds <- ggplot2::diamonds

p1 <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Diamond Price vs Carat",
       x = "Carat",
       y = "Price") +
  theme_minimal()

p2 <- ggplot(diamonds, aes(x = cut, y = price, fill = cut)) +
  geom_boxplot() +
  labs(title = "Diamond Price by Cut Quality",
       x = "Cut",
       y = "Price") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

The correlation plot shows that all variables has strong positive correlation with rspect to price except depth and table.

```{r}
# Select numeric variables
numeric_vars <- diamonds %>%
  select(price, carat, depth, table, x, y, z)

# Compute correlation matrix
cor_matrix <- cor(numeric_vars)

# Plot heatmap
corrplot::corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.8, addCoef.col = "black")
```


The model is fitted with carat, cut, clarity and color as independent variables to predict the price. The model shows that the R square is 91.59%.

```{r}
# Using carat (continuous) + cut + clarity + color
model <- lm(price ~ carat + cut + clarity + color, 
            data = diamonds)

# View model summary
summary(model)
```


Checking the assumptions of the model shows that the assumption of linearity and normality are violated.

```{r}
# Check model diagnostics
par(mfrow = c(2,2))
plot(model)
```

This is the model with only numeric columns as the predictors. The model shows that except "z", all predictors are statistically significant with R square of 85.92% for the model.

```{r}
model_numeric <- lm(price ~ carat + depth + table + x + y + z, 
                    data = diamonds)

summary(model_numeric)

```

